\pagestyle{plain}
\addcontentsline{toc}{chapter}{Acknowledgements}
\chapter*{Acknowledgements}

\lettrine[lraise=0.0, nindent=0.1em, slope=1em]{T}{ his} was a multidisciplinary PhD, and consequently I have many, many people to thank for making this thesis a reality, both formal and informal.

Formally, I'd like to acknowledge the Molecular Organisation and Assembly in Cells (MOAC) Centre for Doctoral Training and the Engineering and Physical Sciences Research Council for giving me the environment and funding to complete my studentship under grant EP/F500378/1. MOAC has been a thoroughly enjoyable experience, and simultaneously, quite possibly the hardest thing I've ever done. I cannot recommend the CDT model strongly enough for anyone that wants to undertake a PhD. On a more individual basis, thanks go to Prof. Alison Rodger, Dr. Hugo van Den Berg and Dr. Nikola Chmel for giving me the opportunity to join, quite frankly, one hell of a community. Thank you also to the indispensable Mrs. Naomi Grew, for putting up with my (and no doubt everyone else's) incessant requirements!

Thanks also go to everyone within the CDT, especially the 2013 cohort, for making the last 4 years so enjoyable. We've had some excellent times at away days and residential courses, and an abundance of stories to tell, whether that be from walking for an hour and a half across London on our way back from a night out in Soho, to watching one another fall in the river Severn whilst canoeing, and a great many more that are probably best left out of a thesis!

Next, I am, of course, indebted to my supervisors, Dr. Nick Waterfield and Prof. Matt Gibson. It has been an absolute pleasure to work with two of the most enthusiastic and insightful scientists I've ever met. Every meeting left another dozen PhD's worth of ideas and work to be done, and this enthusiasm was infectious. This attitude to discovery and science in general is everything I think of when I picture an exemplary scientist. More specifically, thanks to Nick for putting together the project for a lowly Masters student who turned up his office one day, and allowing me to `own' the project from the outset - and for letting me run off around the world at various points attempting to turn his ideas in to a company! On the topic of turning up out of the blue, thanks also go to Matt for agreeing to supervise an unusual project that was somewhat outside the scope of the lab's research at the time, and for really throwing him self into the project. I hope I get to continue this project with you both far in to the future in some form or another.\footnote{Not to mention, they both make excellent drinking companions!}

Additionally I'd like to acknowledge everyone in both of my labs, past and present for providing advice, help, and the camaraderie required to get through a PhD! Particularly, thanks go to Dr. Alexia Hapeshi for all her help throughout my Masters and PhD, a constant source of excellent advice and support in the lab.

Though it doesn't get much `air-time' in this thesis, a journey as exciting as my PhD unfolded along side it, which was my foray in to the world of biotechnology and commercialisation of research. For their encouragement, advice, and support in this process, I have to thank Dr. Laura Lane and Dr. James Lapworth from Warwick Ventures. I hope we get to continue our world tour of jazz bars for a good while to come yet!

In no particular order, I'd like to acknowledge Dr. Andy Millard for his support with bioinformatics tasks, and for agreeing to participate in my advisory panel. Similarly, thanks to Dr. Chris Corre for being on the advisory panel and helping out with submission of projects to the Warwick Synthetic Biology centre; and once again to Dr. Nikola Chmel for also participating in the advisory panel and helping out with Circular Dichroism experiments. I would like to acknowledge the Medical Research Council's Cloud Infrastructure for Microbial Bioinformatics, particularly Marius Bakke, for allowing me access to computer resources that `proper' computational scientists would kill for! I also wish to thank Chris Thoroughgood, Dr. Avinash Punekar and Prof. David Roper for help and access to equipment for protein purification and crystallography experiments.

Last but by no means least, I'd like to thank three of my mentors from the past. I feel like this doesn't happen as often as it should in many acknowledgement sections, and I intend to rectify that - in fact, before this thesis was so much as a scribble on a page or a byte on my laptop, I knew this section would be written. I can categorically say, and without any sense of exaggeration, that I would not have gotten this far without their influences, and they are entirely the reason I chose to pursue the biological sciences. Firstly, Mrs. Elena Segalini-Bower, my GCSE and A Level Biology teacher encouraged me to apply to university and set me on the path of the biological sciences in the very beginning. Without her, I would never have even made it to undergraduate! She is a fabulous teacher (and in my opinion was absolutely robbed of the Royal Society of Biology teacher of the year award in July 2015!), and has no doubt influenced many students since me to come this far! She challenged me at every opportunity as a student, and even if I was less than thrilled about it at the time (I wanted easy homework too!), it has been the absolute making of me as a student and scientist. I really can't express my gratitude enough in words, so I won't labour the point further!

For all the same reasons, Dr. Justin Pachebat and Prof. Luis Mur, my undergraduate tutor/supervisors were instrumental in my understanding and appreciation of the biological sciences. If Mrs. Bower is the reason I made it to undergraduate, there is no doubt in my mind that Justin and Luis are the reason I made it to postgraduate. They pushed me academically, and provided me with incredible freedom and opportunities for my dissertation project which were unequivocally the catalyst for me becoming hooked on experimental biological research, and they supported me every step of the way.

Thank you once again to everyone who has influenced my journey, I wish I had the space to name everyone individually!

\vfill
%\settowidth\longest{\large \itshape This field is not quite the same as others in that it will not tell us much [...].}
\begin{center}

% I am immensely fond of you all, and that eleventy-one years is too short a time to live among such excellent and admirable hobbits.
% I don't know half of you half as well as I should like; and I like less than half of you half as well as you deserve.

\hrule
\vspace{0.5cm}
\parbox{0.8\textwidth}{
\noindent\Ttelco\TTdot\Ts\Ttelco\TTthreedots\Tmalta\Ts\Ttelco\TTdot\Tmalta\TTdoubler\TTacute\Tnuumen\Tsilmenuquerna\TTacute\Tlambe\TTtwodotsbelow\Ts\Tformen\TTrightcurl\Tando\Ts\Ttelco\TTrightcurl\Tformen\Ts\Tanna\TTtwodotsbelow\TTrightcurl\Ttelco\TTleftcurl\Ts\Ttelco\TTthreedots\Tlambe\TTdoubler\Ts\Tcentereddot\Ts\Ttelco\TTthreedots\Tando\Ts\Ttinco\Thyarmen\TTthreedots\Ttinco\Ts\Ttelco\TTacute\Tlambe\TTacute\Tvala\TTacute\Tanto\TTtwodotsbelow\Ttelco\TTrightcurl\Tnuumen\TTacute\Ts\Tanna\TTtwodotsbelow\TTacute\Ttelco\TTthreedots\Toore\Tsilme\Ts\Ttelco\TTdot\Tsilme\Ts\Ttinco\TTrightcurl\Ttelco\TTrightcurl\Ts\Tsilme\Thyarmen\TTrightcurl\Toore\Ttinco\Ts\Ttelco\TTthreedots\Ts\Ttinco\TTdot\Tmalta\TTacute\Ts\Ttinco\TTrightcurl\Ts\Tlambe\TTdot\Tvala\TTacute\Ts\Ttelco\TTthreedots\Tmalta\TTrightcurl\Tanga\Ts\Tsilmenuquerna\TTleftcurl\Tcalma\Thyarmen\Ts\Ttelco\TTacute\Tcalma\Tsilme\Tcalma\TTacute\Tlambe\TTdoubler\TTacute\Tanto\Ts\Ttelco\TTthreedots\Tando\Ts\Ttelco\TTthreedots\Tmalta\TTdot\Troomen\Tlambe\TTacute\Ts\Thyarmen\TTrightcurl\TTdot\Ttinco\Tsilme\Ts\Tcolon\Ts\Ttelco\TTdot\Ts\TTrightcurl\Tnuumen\Ttinco\Ts\Tcalma\Tnuumen\TTrightcurl\Tvilya\Ts\Thyarmen\TTthreedots\Tlambe\Tformen\Ts\Ttelco\TTrightcurl\Tformen\Ts\Tanna\TTtwodotsbelow\TTrightcurl\Ttelco\TTleftcurl\Ts\Thyarmen\TTthreedots\Tlambe\Tformen\Ts\Ttelco\TTthreedots\Tsilme\Ts\Tvilya\TTacute\Tlambe\TTdoubler\Ts\Ttelco\TTthreedots\Tsilme\Ts\Ttelco\TTdot\Ts\Tsilme\Thyarmen\TTrightcurl\Ttelco\TTleftcurl\Talda\Ts\Tlambe\TTdot\Tcalma\TTacute\Ts\Tcentereddot\Ts\Ttelco\TTthreedots\Tando\Ts\Ttelco\TTdot\Ts\Tlambe\TTdot\Tcalma\TTacute\Ts\Tlambe\TTacute\Tesse\Ts\Ttinco\Thyarmen\TTthreedots\Tnuumen\Ts\Thyarmen\TTthreedots\Tlambe\Tformen\Ts\Ttelco\TTrightcurl\Tformen\Ts\Tanna\TTtwodotsbelow\TTrightcurl\Ttelco\TTleftcurl\Ts\Thyarmen\TTthreedots\Tlambe\Tformen\Ts\Ttelco\TTthreedots\Tsilme\Ts\Tvilya\TTacute\Tlambe\TTdoubler\Ts\Ttelco\TTthreedots\Tsilme\Ts\Tanna\TTtwodotsbelow\TTrightcurl\Ttelco\TTleftcurl\Ts\TTacute\Tsilmenuquerna\TTacute\Toore\Tvala\TTacute\Ts\Tcolon
}
\vspace{0.5cm}
\hrule
\end{center}

\vspace{2cm}
\clearpage